Exact Asymptotics for Causal Mediation Analysis

### 2024-06-14 notes/Consolidated Notes/Consolidated Notes.tex
Consolidate notes into one document

\documentclass{article}

@@

@@

\title{Review of the Literature on Exact Formulas for Mediation Effects}

@@

@@

I'm just finishing-up my code for the binary response, binary mediator, mixed effects case. My analysis is running, but the output doesn't look right so I need to do some debugging.

@@

\subsubsection{UQ for Group-Level Effects}
\ref{sec:Imai_UQ}

@@

\bibliography{Lit_Review_Bib}

>>>>

\documentclass{report}

@@

\usepackage{tabu,multirow}

@@

\newcommand{\sY}{\mathcal{Y}}

\title{An Update on William's Work}

@@

\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

I have been working on three main areas: a literature review of estimation and UQ for multilevel mediation, implementing the UQ developed by B \& B for the estimated total effect, and expanding their development to address direct and indirect effects. In this document, I attempt to consolidate this disparate work in one place. Unfortunately, since my implementation was written in RMarkdown, I only overview that work here and refer to an external PDF.

In the interest of minimizing the amount of re-writing necessary, I have organized this document into chapters.

\tableofcontents

\chapter{Literature Review}

@@

I'm just finishing-up my code for the binary response, binary mediator, mixed effects case. My analysis is running, but the output doesn't look right, so I need to do some debugging.

@@

\subsubsection{UQ for Group-Level Effects}
\label{sec:Imai_UQ}

@@

\chapter{Implementation}

I don't have a whole lot to say here. I implemented the $\delta$-method based uncertainty quantification developed by B \& B for the total mediation effect. Please see the RMarkdown document and accompanying PDF called ``William Analysis.Rmd'' and ``William-Analysis.pdf'' for discussion and results.

My findings for the fixed-effects models look pretty reasonable. However, the mixed-effects setting is much less nice. I'm still exploring why this is happening. I hope to have more to say here soon.

\chapter{Extension to Other Mediation Effects}

In this chapter, we develop definitions and formulas for the direct and indirect mediation effects to complement the total effect given by B \& B. We follow the work of \citet{Pea12} and the Samoilenko and Lefebvre group (e.g., \citealp{Sam23}).

\section{Expected Nested Counterfactuals}
\label{sec:nest_CFs}

To start, we define the nested counterfactual $Y(x, M(x'))$ as the value that $Y$ would assume when $X$ is set to $x$ and $M$ is set to whatever value it would have assumed if we had set $X$ to $x'$. \citet{Pea12} identifies the expected value of this nested counterfactual with the following expression based on conditional expectations:
%
\begin{align}
    \bE Y(x, M(x')) = \int \bE(Y | M=m, X=x) \bP(M = dm | X = x') \label{eq:med_gen}
\end{align}
%
Focusing now on the case with binary response and binary mediator, \ref{eq:med_gen} becomes
%
\begin{multline}
    \bE Y(x, M(x')) = \bP(Y=1 | M=1, X=x) \bP(M=1 | X=x') + \\\bP(Y=1 | M=0, X=x) \bP(M=0 | X=x') \label{eq:med_bin}
\end{multline}
%
Consider using  logistic regression models for $Y$ and $M$. Write $\mathrm{logit}(\bE(Y | M=m, X=x)) = \beta_0 + \beta_m m + \beta_x x$ and $\mathrm{logit}(\bE(M | X=x)) = \alpha_0 + \alpha_x x$. Then (\ref{eq:med_bin}) becomes
%
\begin{multline}
    \bE Y(x, M(x')) = \frac{1}{1 + \exp(-\beta_0 - \beta_m - \beta_x x)} \cdot \frac{1}{1 + \exp(-\alpha_0 - \alpha_x x')} +\\
    \frac{1}{1 + \exp(-\beta_0 - \beta_x x)} \cdot \frac{1}{1 + \exp(\alpha_0 + \alpha_x x')} \label{eq:med_bin_fix}
\end{multline}
%
Equation (\ref{eq:med_bin_fix}) holds for logistic regression with fixed-effects only. If we instead use  mixed-effects logistic regressions for $Y$ and $M$, then (\ref{eq:med_gen}) and (\ref{eq:med_bin}) still hold, but (\ref{eq:med_bin_fix}) must be modified. For the mixed-effects models, first write $V = (V_0, V_m, V_x) \sim \mathrm{N}(0, \Sigma_V)$ and $U = (U_0, U_x) \sim \mathrm{N}(0, \Sigma_U)$ for the random-effects in our models for $Y$ and $M$ respectively. Next, write $\mathrm{logit}(\bE(Y | V, M=m, X=x)) = (\beta_0 + V_0) + (\beta_m + V_m) m + (\beta_x + V_x) x$ and $\mathrm{logit}(\bE(M |U, X=x)) = (\alpha_0 + U_0) + (\alpha_x + U_x) x$. Returning now to identification of the expected counterfactual for $Y$, we get
%
\begin{align}
    \bE Y(x, M(x')) =& \left[\int \bP(Y=1 |V=v, M=1, X=x) \bP(V = dv)\right] \cdot \\
    & \left[\int \bP(M=1 |U=u, X=x') \bP(U=du)\right] +\\
     & \left[\int \bP(Y=1 |V=v, M=0, X=x) \bP(V = dv)\right] \cdot \\
     & \left[\int \bP(M=0 |U=u, X=x') \bP(U=du)\right]
\end{align}
%
and, in the logistic regression context,
%
\begin{align}
    \bE Y(x, M(x')) =& \left[ \int \frac{\phi(v; 0, \Sigma_V)}{1 + \exp(-(\beta_0 + v_0) - (\beta_m + v_m) - (\beta_x + v_x) x)}  dv \right] \cdot \label{eq:med_bin_ran}\\
    &\left[ \int \frac{\phi(u; 0, \Sigma_U)}{1 + \exp(-(\alpha_0 + u_0) - (\alpha_x + u_x) x')}  du \right] + \nonumber\\
    & \left[\int \frac{\phi(v; 0, \Sigma_V)}{1 + \exp(-(\beta_0 + v_0) - (\beta_x + v_x) x)} dv \right] \cdot \nonumber\\
    & \left[\int \frac{\phi(u; 0, \Sigma_U)}{1 + \exp((\alpha_0 + u_0) + (\alpha_x + u_x) x')} du \right], \nonumber
\end{align}
%
where $\phi(.; \mu, \Sigma)$ is the multivariate normal density with mean $\mu$ and covariance matrix $\Sigma$.

Note that the four integrals in (\ref{eq:med_bin_ran}) are all multivariate, but can be transformed to univariate integrals by suitable changes of variables. To this end, write $\eta = \alpha_0 + \alpha_x x$ and $\zeta = \beta_0 + \beta_x x$ for two linear predictors (note that $\zeta$ does not contain $\beta_m$), and $\gamma^2_\Sigma(c_1, \ldots, c_r) = (c_1, \ldots, c_r) \Sigma (c_1, \ldots, c_r)^T$, where $\Sigma$ is an $r$-by-$r$ covariance matrix. We will generally set $\Sigma = \Sigma_V$ or $\Sigma = \Sigma_U$, in which case we write $\gamma^2_V$ or $\gamma^2_U$ respectively. We now define the function $\psi$ as a template for the four integrals in (\ref{eq:med_bin_ran}). 
%
$$
\psi(\mu, \sigma^2) := \int \frac{\phi(z; 0, 1)}{1 + \exp(-\mu - \sigma z)} dz.
$$
%
Note that $\psi$ is a univariate integral, so we can expect it to be well-approximated by routine numerical quadrature techniques. We now re-write (\ref{eq:med_bin_ran}) in terms of $\psi$ as follows,
%
\begin{align}
    \bE Y(x, M(x')) =& \psi(\zeta + \beta_m, \gamma^2_V(1, 1, x)) \cdot \psi(\eta, \gamma^2_U(1, x)) +\\
    & \psi(\zeta, \gamma^2_V(1, 0, x)) \cdot \psi(-\eta, \gamma^2_U(1, x)). \nonumber 
\end{align}

\section{Mediation Effects}
Denote the expected nested counterfactual defined in (\ref{eq:med_gen}) by $\sY(x, x') = \bE Y(x, M(x'))$. We can define the various mediation effects in terms of expected counterfactuals. Note that mediation effects for a binary outcome are commonly defined on three different scales: risk difference, risk ratio and odds ratio. Table \ref{tab:med_eff_defs} gives all such definitions explicitly. In the rest of this section, we outline the processes of estimation and uncertainty quantification for the mediation effects defined in Table \ref{tab:med_eff_defs}.

%
\begin{table}[ht]
    \centering
    \caption{Definitions of various mediation effects; $x$ and $x'$ denote different values of the exposure.}
    \label{tab:med_eff_defs}
    \begin{tabu}{|[1pt]c|[1pt]c|c|[1pt]}
        \tabucline[1pt]{-}
        \multirow{3}{*}{Risk Difference} & Total Effect & $\sY(x, x) - \sY(x', x')$ \\
        \cline{2-3}
        & Direct Effect & $\sY(x, x') - \sY(x', x')$\\
        \cline{2-3}
        & Indirect Effect & $\sY(x, x) - \sY(x, x')$ \\
        \tabucline[1pt]{-}
        \multirow{3}{*}{Risk Ratio} & Total Effect & $\sY(x, x) / \sY(x', x')$ \\
        \cline{2-3}
        & Direct Effect & $\sY(x, x') / \sY(x', x')$\\
        \cline{2-3}
        & Indirect Effect & $\sY(x, x) / \sY(x, x')$ \\
        \tabucline[1pt]{-}
        \multirow{3}{*}{Odds Ratio} & Total Effect & $\left. \frac{\sY(x, x)}{1 - \sY(x, x)} \right/ \frac{\sY(x', x')}{1 - \sY(x', x')} $\\
        \cline{2-3}
        & Direct Effect & $\left. \frac{\sY(x, x')}{1 - \sY(x, x')} \right/ \frac{\sY(x', x')}{1 - \sY(x', x')} $\\
        \cline{2-3}
        & Indirect Effect & $\left. \frac{\sY(x, x)}{1 - \sY(x, x)} \right/ \frac{\sY(x, x')}{1 - \sY(x, x')} $\\
        \tabucline[1pt]{-}
    \end{tabu}
\end{table}

\subsection{Estimation}

Mediation effects are all defined in terms of the expected nested counterfactuals, $\sY(x, x')$. As such, the estimation of any mediation effect given in Table \ref{tab:med_eff_defs} centers around estimation of $\sY$. To this end, write $\theta$ for all parameters upon which $\sY$ depends. That is, $\theta$ contains both sets of regression coefficients, $\beta_0, \beta_m, \beta_x$ and $\alpha_0, \alpha_x$, as well as both sets of covariance parameters. For consistency with B \& B, we parameterize these as $\tau_0, \tau_m, \tau_x$ for the standard deviations of $V_0, V_m, V_x$, and $\tau_{0,m}, \tau_{0,x}, \tau_{m,x}$ for the corresponding correlations\footnote{While our notation doesn't match that given by B \& B, parameterizing in terms of the standard deviations and correlations does. Alternative choices include the variances and covariances, or the unique components of the Cholesky factorizations of $\Sigma_V$ and $\Sigma_U$ \citep{Wan18}.}. Similarly, we use $\sigma_0, \sigma_x$ for the standard deviations of $U_0, U_x$, and $\sigma_{0,x}$ for their correlation. We are now equipped to write-out $\theta$ in full. The order of parameters is chosen to match my code (and to avoid refactoring thereof).
$$
\theta = (\alpha_0, \alpha_x, \sigma_0, \sigma_x, \sigma_{0,x}, \beta_0, \beta_m, \beta_x, \tau_0, \tau_m, \tau_x, \tau_{0,m}, \tau_{0,x}, \tau_{m,x})
$$
%
In order to estimate $\theta$, we fit two mixed-effects logistic regression models. The first model predicts the mediator, $M$, using the exposure, $X$, while the second predicts the outcome, $Y$, using $M$ and $X$. Both models also contain intercepts. We include random effects for all regression coefficients in both models. \textcolor{red}{Either or both models may also contain one or more confounders, $\mathbf{C}$, although we assign these confounders only fixed effects.} Following the algorithm used by the \texttt{lme4} package in \texttt{R} \citep{Bat15, Wal23}, we estimate the parameters in both regression models by maximizing an approximation to the marginal likelihood of the response ($M$ or $Y$). This approximation is obtained using Laplace's Method, with a penalized iteratively re-weighted least squares algorithm maximizing to maximize the joint likelihood of the observed data and the random effects. Applying this procedure to both of our regression models yields estimates for all components of $\theta$.

Combining our estimate of $\theta$ with Equation \ref{eq:med_bin_ran} from Section \ref{sec:nest_CFs} and the formulas given in Table \ref{tab:med_eff_defs}, we are now able to estimate all mediation effects on the three scales we consider. It remains however, to address uncertainty quantification.

\subsection{Uncertainty Quantification}

As mentioned above, all mediation effects are defined in terms of expected nested counterfactuals $\sY$, which themselves depend on the parameter vector $\theta$. Quantification of uncertainty for our estimator of a mediation effect thus proceeds in three stages. First, we obtain a covariance matrix for our estimator of $\theta$. Second, we apply the $\delta$-method to get a joint covariance matrix for all values of $\sY$ required to evaluate the mediation effect of interest. Finally, we use the $\delta$-method again to translate uncertainty in the $\sY$s to uncertainty in the mediation effect. In fact, we can obtain the covariance matrix for all three mediation effects defined on a particular scale (e.g., risk ratio), or, indeed, between all nine mediation effects given in Table \ref{tab:med_eff_defs}. Note that the variances and covariances discussed above are asymptotic.

We now proceed with the three steps discussed above, starting with an asymptotic covariance matrix for $\theta$. \textcolor{red}{For now, we treat the two regression models as independent. That is, we assume that the covariance between all parameter estimators in our $M$ model and all parameter estimators in our $Y$ model are zero. This is likely not true; call it a working assumption. See \citet{Bau06} for a method to model this inter-model dependence.} This step is fairly straightforward, since both regression models are fit using (approximate) maximum likelihood. We simply compute the negative Hessian of the marginal log-likelihood for the observed data in both models, and stack the results into a block-diagonal matrix. Call the resulting covariance matrix $\hat{\Sigma}_\theta$. This computation is performed by the \texttt{merDeriv} package in \texttt{R} \citep{Wan18}.

To quantify uncertainty in $\sY$ based on our estimator for $\theta$, we use the $\delta$-method \citep[see, e.g., Chapter 3 of][]{vdV98}. Briefly, given the asymptotic covariance of our estimator for $\theta$, $\sqrt{n}(\hat{\theta} - \theta) \rightsquigarrow \mathrm{N}(0, \Sigma)$, the asymptotic covariance of some function of our estimator can be obtained by differentiating that function and multiplying by the limiting distribution. That is, $\sqrt{n}[f(\hat{\theta}) - f(\theta)] \rightsquigarrow \nabla f(\theta)\mathrm{N}(0, \Sigma) \overset{d}{=} \mathrm{N}(0, \nabla f(\theta) \Sigma \nabla f(\theta)^T)$. We estimate this limiting covariance by $\nabla f(\hat{\theta}) \hat{\Sigma} \nabla f(\hat{\theta})^T$, where typically $\Sigma = \Sigma(\theta)$ and $\hat{\Sigma} = \Sigma(\hat{\theta})$.

Returning now to our problem, in order to apply the $\delta$-method we need the gradient of $\sY$ with respect to $\theta$. While tedious, this calculation is easily performed with the help of symbolic differentiation software like Maple \citep{Map20}. Next, we evaluate our gradient formula at $\hat{\theta}$, $\left. \nabla \sY \right|_{\theta = \hat{\theta}}$, then pre- and post-multiply $\hat{\Sigma}_\theta$ by this estimated gradient. If we require the covariance matrix for multiple values of $\sY$ (e.g., $\sY(x,x)$ and $\sY(x,x')$ for some $x \neq x'$), we frame the problem as a vector-valued transformation, and compute the Jacobian by stacking the gradient for each individual $\sY$. More precisely, to get a covariance matrix for $\sY_1, \ldots, \sY_r$, we first construct the $r$-by-$|\theta|$ Jacobian matrix $\mathcal{J} := [\nabla \sY_1, \ldots, \nabla \sY_r]$. We then evaluate $\mathcal{J}$ at $\theta = \hat{\theta}$, call the result $\hat{\mathcal{J}}$, and pre- and post-multiply $\hat{\Sigma}_\theta$ by $\hat{\mathcal{J}}$, giving $\hat{\mathcal{J}} \hat{\Sigma}_\theta \hat{\mathcal{J}}^T$.

Finally, to get the asymptotic variance of one or more estimated mediation effects, we just apply the $\delta$-method to the corresponding formulas from Table \ref{tab:med_eff_defs}. Since the mediation effects are all simple functions of the $\sY$, any gradients required for this step are easily obtained.

\section{To Do}
\begin{itemize}
    \item Incorporate an interaction term between $X$ and $M$ in the model for $Y$.
    \item Explore dependence between the models for $M$ and $Y$ using the `stacking' technique described in \citet{Bau06}. This will give non-zero covariance between parameter estimators from the two models.
    \item Incorporate covariates/confounders
\end{itemize}

\bibliography{../Literature_Review/Lit_Review_Bib}

### 2024-06-14 notes/William Analysis.Rmd
Consolidate notes into one document

# Number of clusters
all_Ks = c(5, 20, 50)
# K=5

@@

@@

set.seed(2222)

@@

for(j in seq_len(nrow(N_K_combs))){
  n = N_K_combs[j, "N"]
  K = N_K_combs[j, "K"]

  # Containers for output with this value of n
  some_a_hats = list()
  some_theta_hats = list()
  some_b_hats = list()
  some_gamma_hats = list()

  some_M_covs = list()
  some_Y_covs = list()

  for(i in 1:num_reps){
    print(paste0("n = ", n, ", K = ", K, ", i = ", i, " of ", num_reps))

    # Use tryCatch to omit any datasets leading to convergence issues
    tryCatch({
      all_Xs = list()
      all_Ws = list()

@@

      some_a_hats[[i]] = a_hat
      some_theta_hats[[i]] = theta_hat
      some_M_covs[[i]] = M_cov

@@

      some_b_hats[[i]] = b_hat
      some_gamma_hats[[i]] = gamma_hat
      some_Y_covs[[i]] = Y_cov

@@

      ## Mediation effect
      ### See Helpers.R for the function Phi, which computes the mediation effect on odds-ratio scale
      med_hat = Phi(eta_hat, zeta_hat, a_x_hat, b_m_hat, b_x_hat, sigma_M2, sigma_Y2, sigma_M1, sigma_Y1)
      some_med_hats[[i]] = med_hat

@@

      some_med_SEs[[i]] = med_SE
    }, error = function(e){})
  }

  # Store output for current value of n
  all_a_hats[[j]] = some_a_hats
  all_theta_hats[[j]] = some_theta_hats
  all_b_hats[[j]] = some_b_hats
  all_gamma_hats[[j]] = some_gamma_hats

  all_M_covs[[j]] = some_M_covs
  all_Y_covs[[j]] = some_Y_covs

  all_med_hats[[j]] = some_med_hats
  all_med_SEs[[j]] = some_med_SEs
}
toc()

# Save output
# save(all_a_hats, all_theta_hats, all_b_hats, all_gamma_hats, all_M_covs, all_Y_covs, all_med_hats, all_med_SEs, file = "William-Analysis_cache/MC-bin-bin-mix-n=200,500,1000.RData")
# load("William-Analysis_cache/MC-bin-bin-mix.RData")
```

@@

results$Shapiro_p = all_norm_p_vals
results
```

>>>>

# Number of clusters
all_Ks = c(5, 20, 50, 100)
# K=5

@@

library(doParallel)
library(pbapply)

@@

set.seed(1)

@@

# Initialize cluster on my machine ----
# n_cores = 2
# n_cores = 10
n_cores = parallel::detectCores() - 1
my_cluster = makeCluster(n_cores, type = "PSOCK")
registerDoParallel(my_cluster)
clusterEvalQ(my_cluster,{
  library(MASS)
  library(lme4)
  library(merDeriv)
  source("../src/Helpers.R")
})
clusterExport(my_cluster, c("N_K_combs", "num_reps", "a_0", "a_1", "A_2", "a", "b_0", "b_1", "b_2", "B_3", "b", "Sigma_a", "theta", "Sigma_b", "gamma", "p_conf", "x_pred", "W_pred"))
clusterSetRNGStream(my_cluster, iseed = 1)

# output_bin_bin_ran = pblapply(seq_len(nrow(N_K_combs)), function(j){
# output_bin_bin_ran = pblapply(nrow(N_K_combs), function(j){
for(j in seq_len(nrow(N_K_combs))){
  n = N_K_combs[j, "N"]
  K = N_K_combs[j, "K"]

  clusterExport(my_cluster, c("n", "K"))

  print(paste0("j = ", j, " of ", nrow(N_K_combs),  " (n=", n, ", K=", K, ")"))

  # # Containers for output with this value of n
  # some_a_hats = list()
  # some_theta_hats = list()
  # some_b_hats = list()
  # some_gamma_hats = list()
  # 
  # some_M_covs = list()
  # some_Y_covs = list()
  # 
  # some_med_hats = list()
  # some_med_SEs = list()

  # for(i in 1:num_reps){
  some_output = pblapply(seq_len(num_reps), function(i){
    # print(paste0("n = ", n, ", K = ", K, ", i = ", i, " of ", num_reps))

    this_output = NULL

    # Use tryCatch to omit any datasets leading to convergence issues
    tryCatch({                                                                    ############## Un-quote
      all_Xs = list()
      all_Ws = list()

@@

      # some_a_hats[[i]] = a_hat
      # some_theta_hats[[i]] = theta_hat
      # some_M_covs[[i]] = M_cov

@@

      # some_b_hats[[i]] = b_hat
      # some_gamma_hats[[i]] = gamma_hat
      # some_Y_covs[[i]] = Y_cov

@@

      ## Mediation effect
      ### See Helpers.R for the function Phi, which computes the mediation effect on odds-ratio scale
      med_hat = Phi(eta_hat, zeta_hat, a_x_hat, b_m_hat, b_x_hat, sigma_M2, sigma_Y2, sigma_M1, sigma_Y1)
      # some_med_hats[[i]] = med_hat

@@

      # some_med_SEs[[i]] = med_SE

      this_output = list(a_hat = a_hat, theta_hat = theta_hat, b_hat = b_hat, gamma_hat = gamma_hat, M_cov = M_cov, Y_cov = Y_cov, med_hat = med_hat, med_SE = med_SE)
    # }, error = function(e){grad_Phi_obs <<- e}) # For troubleshooting parallel errors
    }, error = function(e){})                                                       ############ Un-quote

    return(this_output)

  }, cl = my_cluster)
  # })

  # Store output for current values of n and K
  all_a_hats[[j]] = lapply(some_output, function(x) x$a_hat)
  all_theta_hats[[j]] = lapply(some_output, function(x) x$theta_hat)
  all_b_hats[[j]] = lapply(some_output, function(x) x$b_hat)
  all_gamma_hats[[j]] = lapply(some_output, function(x) x$gamma_hat)
  all_M_covs[[j]] = lapply(some_output, function(x) x$M_cov)
  all_Y_covs[[j]] = lapply(some_output, function(x) x$Y_cov)
  all_med_hats[[j]] = lapply(some_output, function(x) x$med_hat)
  all_med_SEs[[j]] = lapply(some_output, function(x) x$med_SE)
}

runtime = toc()

parallel::stopCluster(my_cluster)

# Save output
# save(all_a_hats, all_theta_hats, all_b_hats, all_gamma_hats, all_M_covs, all_Y_covs, all_med_hats, all_med_SEs, runtime, file = "William-Analysis_cache/MC-bin-bin-mix.RData")
# load("William-Analysis_cache/MC-bin-bin-mix.RData")
```

@@

results$Shapiro_p = all_norm_p_vals
results
